---
title: "Magrittr"
---

## Magrittr
`{magrittr}` is a package that allows us to perform more operations within a timeline to make our code more efficient to write and to read. Its main implementation is the pipe (`%>%`) which we already discussed in [Dplyr](dplyr.qmd). However, it also offers as a number of aliases to make operators available in our pipelines.

## . and `
Before we get into the aliases offered by `{magrittr}`, we should realize that we can do without these aliases by using dots (.) and backticks (`). Here, the dot indicates the data in its current form (i.e. after all previous operations have been applied) and the backticks are put around the operator (e.g. \grave+\grave).

For instance, take the below data frame:

```{r}
#| code-fold: false
# Create example data
dat_fib <- tibble(id = 1:10,
                  value = c(1, 1, 2, 3, 5, 8, 13, 21, 34, 55))

```

If we want to substract 3 from all numbers and then multiply by 1.5, we could do the following:

```{r}
#| code-fold: false
# First, load magrittr
pacman::p_load("magrittr")

# Start pipeline with vector
dat_fib %>%
    # Substract 3
    .`-` 3 %>%
    # Multiply by 1.5
    .`*` 1.5

```


## Aliases

## |>